Comparing probabilistic predictive models applied to football

### 2015-08-21 template.tex
Ajeitando alteracoes.

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{We create three simple multinomial-Dirichlet models and compare them to  two 
well-known sophisticated statistical models for association football (soccer) predictions according to their performance in predicting the full-time results for the second round of the 2014 Brazilian first division championship. The predictive models are compared using three proper scoring rules, the proportion of errors and the calibration assessment. Our results show similar predictive performances for all the five methods, thus {\color{blue} indicating that the new models offer simpler and more interpretable
predictions while maintaining predictive power. This  discourages the use of the two popular models for match result forecast in the Brazilian football championship.}}

@@

	Bothered by such concerns, our goal was to answer the following question: is it possible, using a simple probabilistic model, to predict results of football matches better than more sophisticated or complex models?
	{\color{blue}To answer this, we developed simple multinomial-Dirichlet models that consider only the number of matches won, tied or lost by each team as inputs. We then compared them with the two most popular 
	 models used to predict the outcomes of the Brazilian football championship. These two sophisticated models are widely used by football fans, and  they consider multiple covariates as inputs.}
	The models were tested for the second round of the Brazilian football championship of first division and compared using standard metrics or scoring rules. 
	We also used other criteria such as the proportion of matches that were predicted ``correctly'' by each model and a measure of calibration, both explained in detail below.
	According to the adopted criteria, the simple models showed a similar predictive performance to the complex ones.

	The predictions of the two benchmark models are published on Internet websites before every matchday, and are treated by us as black-box models (BB1 and BB2).
	Since one of the authors is responsible for one of these models and the other was described in a dissertation \citep{arruda2000}, we know that both models are based on Holgate bivariate Poisson regression considering as explanatory variables the home field advantage, attack and defense strength of each team, and provide the probabilities of all possible outcomes (home team wins, draws or loses) for each match.
{\color{red}sugiro tirar - Because both models are based on elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models. 
	}

	{\color{red}sugiro tirar -To be precise, we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model, which is not exactly the definition we will use, since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.}

@@

	Our goal is to compute the predictive posterior distribution of the
	upcoming match, $X_{n+1}$, that is,
	$P(X_{n+1}=x|M_1=n_1,M_2=n_2,M_3=n_3)$, $x\in\mathcal{X}$. Using
	Bayesian inference this probability can be easily computed as
	follows. Suppose that $\boldsymbol{\theta}$ has Dirichlet prior distribution
	with parameter $(\alpha_1,\alpha_2,\alpha_3)$, denoted
	$\mathcal{D}(\alpha_1,\alpha_2,\alpha_3)$, with density function
	\[
	\pi(\boldsymbol{\theta}|\boldsymbol{\alpha})=\frac{\Gamma(\alpha_1+\alpha_2+\alpha_3)}{\Gamma(\alpha_1)\Gamma(\alpha_2)\Gamma(\alpha_3)}\theta_1^{\alpha_1-1}\theta_2^{\alpha_2-1}(1-\theta_1-\theta_2)^{\alpha_3-1}
	\]
	\noindent for $\alpha_1$, $\alpha_2$, $\alpha_3 > 0$, then the
	posterior distribution of $\boldsymbol{\theta}$ is
	$\mathcal{D}(n_1+\alpha_1,n_2+\alpha_2,n_3+\alpha_3)$. Thus, the
	predictive distribution of $X_{n + 1}$ is given by the
	integral
	$$
	P(X_{n + 1} = x | M_1 = n_1, M_2 = n_2, M_3 = n_3) = \int_{\boldsymbol{\Theta}} P(X_{n
		+ 1} = x | \boldsymbol{\theta}) \pi(\boldsymbol{\theta} | M_1 = n_1, M_2 = n_2, M_3
	= n_3) d\boldsymbol{\theta},
	$$
	which leads to the following probabilities of win, tie and loss:
	\begin{align*}
	P(X_{n+1} = 1 | M_1=n_1,M_2=n_2,M_3=n_3) &=
	\frac{n_1+\alpha_1}{n+\alpha_{\bullet}}\\
	& \\
	P(X_{n+1} = 2 | M_1=n_1,M_2=n_2,M_3=n_3) &=
	\frac{n_2+\alpha_2}{n+\alpha_{\bullet}} \\
	& \\
	P(X_{n+1} = 3 | M_1=n_1, M_2=n_2, M_3=n_3) &=
	\frac{n_3+\alpha_3}{n+\alpha_{\bullet}}
	\end{align*}
	\noindent where $\alpha_{\bullet} =\alpha_1+\alpha_2+\alpha_3$.

	In this work, we are interested in checking the predictive performance of the Multinomial-Dirichlet
	model for the result of every single match in the second round of the championship given the results
	of all the previous matches observed in the same tournament. Predictions
	are carried out by using a Bayes information updating mechanism. First, we use full-time
	results from the first round as historical data for construction of the Dirichlet prior: we assign a uniform prior on
	$\boldsymbol{\theta}$ over the 2-simplex, i.e. $\mathcal{D}(1, 1, 1)$, and combine this prior with
	the first-round data, obtaining a posterior Dirichlet distribution through conjugacy that represents
	the information about $\boldsymbol{\theta}$ up to the first round. Then, the posterior of the first
	round becomes the prior for the second round, which, for every matchday in the second round, will be combined with all the observed second-round matches
	up to that matchday in order to yield posterior predictive distributions. Note that, on the other hand, the two
	black-box models make predictions from a frequentist perspective; BB1 uses scores of every match
	in the previous two years, while BB2 uses the match scores of the current championship only.

@@

@@

	We therefore conclude that all methods have similar predictive power, and none of them is more informative in average.
	Nevertheless, all methods yielded better predictions than the trivial prediction, and they are well-calibrated.

@@

	{\color{blue}The benchmark models used in this work were chosen because of their wide popularity among football fans in Brazil.
	Several other models are also available in the literature, some of which are even more complex.
	Among them, we can cite those based on modeling the match as a stochastic process evolving through time \citep{Dixon98, Volf2009, Titman2015}, those allowing for the team performance parameters to change along the season \citep{Rue2000, Crowder2002, Owen2011, Koopman2015} and those modeling dependence between number of goals by means of bivariate count distributions \citep{Dixon97, Karlis2003, McHale2007, McHale2011}. 
Contrary to the simple multinomial models we propose,	some of these approaches are able to answer several questions;  for instance, they can estimate teams' performance parameters allowing for ranking teams according to attack and deffensive quality, and 
 they can also predict the final outcomes of matches. We therefore do not claim that our models should replace
	current state-of-the-art approaches. Instead, this work
	indicates that while there has been a boom in the literature with the proposal of several fancy and involved models,
	simple models can many times predict the final outcome as well as their complex counterparts. 
	The importance of such finding is directly related to the model's usability in practice: creators of the
	benchmark models used in this work often say a difficulty they face is that
	wrong predictions often  generate anger in football fans, which is then translated into distrust
	in subsequent predictions. Because of the complexity of those models, they find it hard to explain predicted outcomes
	to users. This is where using simple models pays off:  the multinomial models yield results that are
	easy to explain; they simply involve counts of losses, wins and draws!
	They allow us to offer simple explanations to the football fans about the predictions that were made.}

{\color{blue}The fact that simple models are often as effective as their competitors is often unnoticed by users of such methods, which leads to the overuse of models that are often
	hard to interpret and to implement, and that also require one to keep track of several covariates. We would call this phenomenon the {\it king's nudity}  in allusion to Anderson's {\it The Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''. In this spirit, since we have explored only the prediction of football matches in the specific case o Brazilian football championship, we would like to study if the same extends to other fields such as, for instance, weather or economics forecasts.}

		If, on the one hand, our findings appear as another supportive example of Ockham's razor, on the other they pose several questions about probabilistic prediction of sport events.
		In particular, based on the fact that the models have similar performance, one may ask: is there an irreducible ``randomness'' or ``degree of unpredictability'' implicit in these events?
		Is this degree an indicator of how tight or leveled is the championship being studied?

@@

@@

			\bibitem{emperor}
			Andersen, H.  C. The emperor's new clothes. Genesis Publishing Pvt Ltd, 1987.

@@

@@

			\bibitem{lad}
			Lad, F. {\it Operational Subjective Statistical Methods: a mathematical,
				philosophical, and historical introduction}. New York: Wiley, 1996.

>>>>

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{We create three simple multinomial-Dirichlet models and compare them to  two 
well-known sophisticated statistical models for association football (soccer) predictions according to their performance in predicting the full-time results for the second round of the 2014 Brazilian first division championship. The predictive models are compared using three proper scoring rules, the proportion of errors and the calibration assessment. Our results show similar predictive performances for all the five methods, thus {\color{blue} indicating that the new models offer simpler and more interpretable
predictions while maintaining predictive power.}}% This  discourages the use of two popular models for match result forecast in the Brazilian football championship.}}

@@

	Bothered by such concerns, our goal was to answer the following question: is it possible, using a simple probabilistic model, to predict results of football matches better than more sophisticated or complex models?
	{\color{blue}To answer this, we developed simple multinomial-Dirichlet models that consider only the number of matches won, tied or lost by each team as inputs. We compared these models with two popular models in Brazil, whose predictions for matches of the Brazilian national championships are published on Internet websites. These two sophisticated models are widely consulted by football fans and consider multiple covariates as inputs.}
	The models were tested for the second round of the Brazilian football championship of first division and compared using standard metrics or scoring rules. 
	We also used other criteria such as the proportion of matches that were predicted ``correctly'' by each model and a measure of calibration, both explained in detail below.
	According to the adopted criteria, the simple models showed a similar predictive performance to the complex ones.

	The predictions of the two benchmark models are published on Internet websites before every matchday, and are treated by us as black-box models (BB1 and BB2).
	Since one of the authors is responsible for one of these models and the other was described in a dissertation \citep{arruda2000}, we know that both models are based on Holgate bivariate Poisson regression considering as explanatory variables the home field advantage, attack and defense strength of each team, and provide the probabilities of all possible outcomes (home team wins, draws or loses) for each match.
%{\color{red}sugiro tirar - Because both models are based on elaborate parametric assumptions, they are suitable for the comparison we are proposing: simple versus complex models.}

	%{\color{red}sugiro tirar -To be precise, we would have to define what is a ``simple'' model and what is a ``complex'' model. Based on the opening quotations, one may say that the number of parameters defines the complexity of a model, which is not exactly the definition we will use, since we consider other criteria, like the interpretation of explanatory variables and mathematical structure of the statistical model.}

@@

	Our goal is to compute the predictive posterior distribution of the
	upcoming match, $X_{n+1}$, that is,
	$P(X_{n+1}=x|M_1=n_1,M_2=n_2,M_3=n_3)$, $x\in\mathcal{X}$. Using
	Bayesian inference this probability can be easily computed as
	follows. Suppose that $\boldsymbol{\theta}$ has Dirichlet prior distribution
	with parameter $(\alpha_1,\alpha_2,\alpha_3)$, denoted
	$\mathcal{D}(\alpha_1,\alpha_2,\alpha_3)$, with density function
	\[
	\pi(\boldsymbol{\theta}|\boldsymbol{\alpha})=\frac{\Gamma(\alpha_1+\alpha_2+\alpha_3)}{\Gamma(\alpha_1)\Gamma(\alpha_2)\Gamma(\alpha_3)}\theta_1^{\alpha_1-1}\theta_2^{\alpha_2-1}(1-\theta_1-\theta_2)^{\alpha_3-1}
	\]
	\noindent for $\alpha_1$, $\alpha_2$, $\alpha_3 > 0$, then the
	posterior distribution of $\boldsymbol{\theta}$ is
	$\mathcal{D}(n_1+\alpha_1,n_2+\alpha_2,n_3+\alpha_3)$. Thus, the
	predictive distribution of $X_{n + 1}$ is given by the
	integral
	$$
	P(X_{n + 1} = x | M_1 = n_1, M_2 = n_2, M_3 = n_3) = \int_{\boldsymbol{\Theta}} P(X_{n
		+ 1} = x | \boldsymbol{\theta}) \pi(\boldsymbol{\theta} | M_1 = n_1, M_2 = n_2, M_3
	= n_3) d\boldsymbol{\theta},
	$$
	which leads to the following probabilities of win, tie and loss:
	\begin{align*}
	P(X_{n+1} = 1 | M_1=n_1,M_2=n_2,M_3=n_3) &=
	\frac{n_1+\alpha_1}{n+\alpha_{\bullet}}\\
	& \\
	P(X_{n+1} = 2 | M_1=n_1,M_2=n_2,M_3=n_3) &=
	\frac{n_2+\alpha_2}{n+\alpha_{\bullet}} \\
	& \\
	P(X_{n+1} = 3 | M_1=n_1, M_2=n_2, M_3=n_3) &=
	\frac{n_3+\alpha_3}{n+\alpha_{\bullet}}
	\end{align*}
	\noindent where $\alpha_{\bullet} =\alpha_1+\alpha_2+\alpha_3$.
	{\color{darkgreen} In fact, the multinomial-Dirichlet is a classical model used in several applied works and more information about it can be found in \cite{good1965}, \cite{bernardo1994} and references therein.}

	In this work, we are interested in checking the predictive performance of the Multinomial-Dirichlet
	model for the result of every single match in the second round of the championship given the results
	of all the previous matches observed in the same tournament. Predictions
	are carried out by using a Bayes information updating mechanism. First, we use full-time
	results from the first round as historical data for construction of the Dirichlet prior: we assign a uniform prior on
	$\boldsymbol{\theta}$ over the 2-simplex, i.e. $\mathcal{D}(1, 1, 1)$, and combine this prior with
	the first-round data, obtaining a posterior Dirichlet distribution through conjugacy that represents
	the information about $\boldsymbol{\theta}$ up to the first round. Then, the posterior of the first
	round becomes the prior for the second round, which, for every matchday in the second round, will be combined with all the observed second-round matches
	up to that matchday in order to yield posterior predictive distributions. {\color{darkgreen}For more on the uniform prior on the simplex, see \cite{agresti2010} and \cite{good1965}.}

	Note that, on the other hand, the two
	black-box models make predictions from a frequentist perspective; BB1 uses scores of every match
	in the previous two years, while BB2 uses the match scores of the current championship only.

@@

	%{\color{darkgreen}Therefore, our analyses show that, marginally, all methods yield similar scores in average, i.e., they have similar predictive power. A related question
	%is whether, for each match, the performance of the methods is also similar. Figure \ref{fig::scores2} displays pairwise comparisons
	%of the various score between the five models we investigate. All methods present
	%a reasonable agreement among each other. Evidently, the Multinomial-Dirichlet models
	%are the ones where there is largest agreement. To formally test the hypotheses that the models yield the same scores in average for
	%each match,
	%we use Friedman's rank sum test. All p-values are not significant at the level 5\% (Brier: 0.077; Spherical: 0.055; Proportion
	%of errors: 0.655), except for the logarithmic
	%score, which has p-value of 0.021. Post-hoc analyses show that the differences are in Mn-Dir$_1$ versus BB2 (p=0.031),
	%Mn-Dir$_2$ versus BB1 (p=0.043), and Mn-Dir$_3$ versus BB1 (p=0.030). These results thus indicate that BB1 and BB2
	%are similar to each other in terms of predictive power, while the performance of Mn-Dir models differ from the BB's \emph{for each match}.
	%We therefore conclude that while the models we investigate may yield different scores for each
	%match according to some metrics, the overall scores are similar. That is, while in some matches Mn-Dir models
	%have better results, for others BB models have better performance. However, in the long term, all methods have similar accuracy.
	%}
	{\color{darkgreen}Therefore, our analyses show that all methods yield similar scores in average, i.e., they have similar predictive power. A related question
	is whether, for each match, the performance of the methods is also similar. To answer this question we compared the models pairwise, and the result is found on Figure \ref{fig::scores2}, where the plots display the scores of each match for every couple of models considered. The plots show that all methods are reasonably similar and, evidently, the Multinomial-Dirichlet models
	are the ones that agree the most. To formally test the hypotheses that the models yield the same scores in average for each match, we used Friedman's rank sum test. Except for the logarithmic score (p-value 0.021), all the other metrics did not present significant differences at the level of confidence of 5\% (Brier: 0.077; Spherical: 0.055; Proportion of errors: 0.655). Post-hoc analyses show that the differences are in Mn-Dir$_1$ versus BB2 (p=0.031),
	Mn-Dir$_2$ versus BB1 (p=0.043), and Mn-Dir$_3$ versus BB1 (p=0.030). These results thus indicate that BB1 and BB2
	are similar to each other in terms of predictive power, while the performance of Mn-Dir models differ from the BB models \emph{for each match}.
	We therefore conclude that while the models we investigate may yield different scores for each
	match according to some metrics, the overall scores are similar, i. e., while in some matches Mn-Dir models
	have better results, for others BB models have better performance. However, for a large set of matches, all methods have similar accuracy.
	}

		\begin{figure}[H]
			\centering
			\includegraphics[page=13,scale=0.48]{futebolComparacaoModelosForPaper.pdf}
			\includegraphics[page=14,scale=0.48]{futebolComparacaoModelosForPaper.pdf}\\
			\caption{Pairwise comparison of the various scores. Left plot: upper right plots display Brier Scores;
			lower left plots display Logarithmic Scores. Right plot: upper right plots display proportion of agreements
			between methods; lower left  plots display Spherical Scores. Red line represents the identity $x=y$.}
			\label{fig::scores2}
		\end{figure}

@@

	{\color{darkgreen}Therefore, we once more highlight that 
	although the accuracy of each method may depend on the specific characteristics of each match,
	all methods have similar predictive power, and none of them is more informative  in average. 
    Moreover, all methods yielded better predictions than the trivial prediction, and they were all well-calibrated.}

@@

	{\color{blue}The benchmark models used in this work were chosen because of their wide popularity among football fans in Brazil.
	Several other models are also available in the literature, some of which are even more complex.
	Among them, we can cite those based on modeling the match as a stochastic process evolving in time \citep{Dixon98, Volf2009, Titman2015}, those allowing for the team performance parameters to change along the season \citep{Rue2000, Crowder2002, Owen2011, Koopman2015} and those modeling dependence between number of goals by means of bivariate count distributions \citep{Dixon97, Karlis2003, McHale2007, McHale2011}. 
Contrary to the simple multinomial models we propose, some of these approaches are able to answer several questions, for instance, they can estimate teams' performance parameters allowing to rank the teams according to their offensive and defensive qualities, and can also predict the number of goals scored of a particular match.

Therefore, we do not claim that our models should replace
	current state-of-the-art approaches. Instead, this work points out that, while there has been a boom in the literature with the proposal of several fancy and involved models,
	simple models can many times predict the final outcome equally well as their complex counterparts. 
	The importance of such finding is directly related to the models usability in practice: professionals that use the mentioned
	benchmark models often say a difficulty they face is that
	wrong predictions often generate anger in football fans, which is then translated into distrust
	in subsequent predictions. Because of the complexity of these models, they find hard to explain to the lay user how the outcomes were predicted.
	This is where using simple models pays off:  the multinomial models yield results that are
	easy to explain and only involve counts of losses, wins and draws, allowing one to offer simple explanations to football fans and journalists about the proposed predictions.}

{\color{blue}The fact that simple models are often as effective as their competitors in average is often unnoticed by users of such methods, which leads to the overuse of models that are often
	hard to interpret and to implement, and also requires one to keep track of several covariates. We would call this phenomenon the {\it king's nudity} in allusion to Anderson's {\it The Emperor's New Clothes}, \cite{emperor}: ``But he isn't wearing anything at all!''. In this spirit, since we have explored only the prediction of football matches in the specific case of the Brazilian championship, we would like to study if the same extends to other fields such as, for instance, weather or economics forecasts.}

		If, on the one hand, our findings appear as another supportive example of Ockham's razor, on the other they pose several questions about probabilistic prediction of sport events.
		In particular, based on the fact that the models have similar {\color{darkgreen}predictive power in average}, one may ask: is there an irreducible ``randomness'' or ``degree of unpredictability'' implicit in these events?
		Is this degree an indicator of how tight or leveled is the championship being studied?

@@

			\bibitem{agresti2010}
			Agresti, A. {\em Analysis of Ordinal Categorical Data}. Wiley, 2010.

@@

			\bibitem{emperor}
			Andersen, H.  C. The emperor's new clothes. Genesis Publishing, 1987.

			\bibitem{bernardo1994}
			Bernardo, J. M. and Smith, A. F. M. {\em Bayesian Theory}. Wiley, 1994.

@@

			\bibitem{good1965}
			Good, I. J. {\em The Estimation of Probabilities: An Essay on Modern Bayesian Methods}. MIT Press, 1965.

@@

			\bibitem{lad}
			Lad, F. {\it Operational Subjective Statistical Methods: a mathematical,
				philosophical, and historical introduction}. Wiley, 1996.
